Ornaments as indicators of social change before and after European arrival at Kiwulan, Northeastern Taiwan

### 2019-11-27 analysis/paper/paper.Rmd
revision based on comments

---
title: "Ornaments as indicators of social change before and after European arrival at Kiwulan, Northeastern Taiwan"
author:
  - Li-Ying Wang
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
bibliography: references.bib
csl: "../templates/antiquity.csl" # Insert path for the bib-style
abstract: |
  Long-lasting indirect impacts on indigenous peoples in the periphery of colonial control are poorly understood, especially in East Asia. Trade ornaments from Kiwulan (1400-1900 AD) in northeastern Taiwan show the indirect impacts of European colonial activities on local societies. The diversity of ornaments was greater during the period of European presence compared to previous periods, and their spatial distribution was more clustered. This hints at increasing social inequality resulting from a colonial influence. Ornaments give insights into the increasing social inequality stimulated by European colonisation, and show the agency of indigenous people to incorporate ornaments into their social system.
keywords: |
  Historical archaeology; Iron Age; European colonisation; Trade ornaments; Exchange networks; Taiwan; Monte Carlo simulation
highlights: |
  1. Evidence of the quantity, variety, and spatial distribution of trade ornaments in both local cultural and colonial context is provided.
  2. Results of hypothesis testing point to greater amount and diversity of ornament types with non-random, and most highly concentrated spatial clusters during the European presence.
  3. Trade ornament as a prestige good may result in an increase in social inequality due to competition of wealth accumulation between individuals.
  4. Trade ornament can be a good proxy to detect indirect colonial influence on local indigenous populations and inform on social changes relating to trade with colonial groups.
---

@@

One of the most commonly traded types of object in this region were personal ornaments such as glass and stone beads [@Chen2007; @Li2014; @NMTH2005]. Personal adornments in the archaeological record are useful as signal of an individual's status [@Joyce2005; @Scaramelli2005]. For example, the consumption of stone beads in Southeast Asia during Iron Age is often associated with increasing social stratification or socio-political complexity [@Bellina2014; @Carter2016; @Francis2002; @Theunissen2000; @Kenoyer2000]. In this paper, I explore archaeological ornaments from Kiwulan spanning the pre-European contact period, the period of Spanish and Dutch presence, and the period of Chinese presence. I address the question of whether indirect colonial influences on the indigenous populations can be detected through the ornament assemblages.

# Cultural context of ornaments in Yilan, northeastern Taiwan

The Spanish founded Fort San Salvador at Keelung in 1626 and Fort San Domingo in 1629 at Tamsui, and sent missionaries to local indigenous settlements in this region for religious purposes [@Blusse2000, pp. 343]. A report by a Dominican priest in 1632 mentioned that the Taparri, an indigenous tribe from northern Taiwan, exchanged agate beads with other indigenous groups. This form of exchange was widespread and even the Spanish soldiers learned to use agate as bargaining chips for gambling [@LiandWu2006, pp. 132-149]. The use of beads as prestige goods is indicated by their role in bride price payments, and compensation to resolve disputes [@LiandWu2006, pp. 132-149]. Other records mention that the female shamans (majuorbol) in the tribe would use agate beads as magical items in ritual practice for body healing [@Borao2009, pp. 122-151]. Records of an indigenous funeral process document the use of agate beads in ritual contexts, with more agate beads, pottery, and cloth placed into the graves of more influential people to indicate a family’s higher prestige [@LiandWu2006, pp. 153]. These historical accounts indicate that agate beads were already treated as prestige goods in Yilan before the arrival of Europeans.

@@

The chronology of Kiwulan can be divided into two phases represented by a upper component (700-1200 AD) and a lower component (1400-1900 AD) separated by a sterile. This chronology is based on the stratigraphy and a series of 32 radiocarbon dates previously published by @Chen2007. This paper focuses on the upper component because this has a sequence spanning pre-European contact, European presence, and Chinese presence. Previous researchers divided the Upper Layer Culture component into six layers spanning from the 14th century to the 19th century with hundred year intervals according to the radiocarbon dates, excavation depth, consistency of contexts, and types of diagnostic porcelains such as blue and white porcelains [@Hsieh2009; @Wang2011]. Due to ambiguities in the previous chronology, I re-examined the excavation records to produce a new chronology for the sampled squares. For each sampled square I separated excavation units into three episodes of time: pre-European period, European period, and Chinese period to indicate the dominant foreign influence and cultural interaction for different time periods.

@@

This study focuses on `r sam_post_hole` ornaments from 40 sampling squares located at the dwelling place of Kiwulan, indicated by aliened postholes with in-situ posts (Figure \@ref(fig:KWL-sam-area)). These postholes were found aligned in a north-south direction in some intervals with construction marks, which were interpreted as the remains of silt house structures. At the north margin of the dwelling place were burials that are mostly oriented in an east-west direction across the sampling area. There are 30 burials in the sampling area account for one thrid of the total number of burials at Kiwulan. Ornaments are commonly used as grave goods in burials, and the total number including different ornament classes are 3,173. The use of ornaments as body adornments in burials will be discussed to understand the functions of ornaments, but not statistically analysed since the number of ornaments is highly affected by the presence of bead strings, which sometimes contains thousands of beads in an individual burial [@Chen2007]. In addition to ornaments from the dwelling place and burial, there are 27 ornaments found in middens, which to some extent supports that ornaments were treated as high value items at Kiwulan because of small number in the midden contexts compared to high frequency in burial contexts. It is important to analyse the detailed distribution of ornaments from the dwelling place as it can provide key information concerning the spatial pattens across the space as well as changes in ornament relating to social organisation across different time periods.

The ornaments found in the  Figure \@ref(fig:plot-all-ornaments-count) shows that the most common ornament are golden beads (n = `r golden_bead_n`), followed by glass beads (n =  `r glass_bead_n`), agate beads (n = `r agate_bead_n`), metal rings (n = `r metal_ring_n`), and metal bells (n = `r metal_bell_n`, see also Figure \@ref(fig:ornaments-image)). The variety of subtypes and their individual frequencies are summarized in Table \@ref(tab:tidy-data-for-table).

@@

Figure \@ref(fig:plot-all-densities) shows that there is one major core area during the pre-European period, multiple core areas during European period, and a single core during the Chinese period. There are three consistent sub-regions with a core area that shifts over time. The distribution might indicate increase and decrease in the number of different household-based social groups who possessed more ornaments. The multiple groups during the European period might reflect more unequal consumption of ornaments across the site, relative to other periods, or random patterns resulting from bigger sample size. In addition, the generation of core areas might be biased due to small sample sizes, for example, a few ornaments found at one single square during the Chinese period could create an obvious hotspot.

@@

An indirect colonial presence is indicated at Kiwulan by a greater diversity of ornaments types and materials during the European period. This likely results from participation in large scale exchange networks that stimulated the circulation of different ornament classes. However, the frequency of overall ornaments and each subtype declines significantly during the Chinese period. This may be due to a smaller scale of trading network, the overall decline of the indigenous population in Yilan, and adopting the practices of Han Chinese who did not use these kinds of beads in their dress. Since the end of the 18th century, many indigenous people moved southwards to Hualien due to population pressure caused by Han Chinese immigrants [@Chen2007].

>>>>

---
title: "Ornaments as indicators of social change before and after European arrival at Kiwulan, Northeastern Taiwan"
author:
  - Li-Ying Wang
  - Ben Marwick
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
bibliography: references.bib
csl: "../templates/antiquity.csl" # Insert path for the bib-style
abstract: |
  Long-lasting indirect impacts on indigenous peoples in the periphery of colonial control are poorly understood, especially in East Asia. Trade ornaments from Kiwulan (1400-1900 AD) in northeastern Taiwan show the indirect impacts of European colonial activities on local societies. The diversity of ornaments was greater during the period of European presence compared to previous periods, and their spatial distribution was more clustered. This hints at increasing social inequality resulting from a colonial influence. Ornaments give insights into the increasing social inequality stimulated by European colonisation, and show the agency of indigenous people to incorporate ornaments into their social system.
keywords: |
  Historical archaeology; Iron Age; European colonisation; Trade ornaments; Exchange networks; Taiwan; Monte Carlo simulation
highlights: |
  1. Evidence of the quantity, variety, and spatial distribution of trade ornaments in both local cultural and colonial context is provided.
  2. Results of hypothesis testing point to greater amount and diversity of ornament types with non-random, and most highly concentrated spatial clusters during the European presence.
  3. Trade ornament as a prestige good may result in an increase in social inequality due to competition of wealth accumulation between individuals.
  4. Trade ornament can be a good proxy to detect indirect colonial influence on local indigenous populations and inform on social changes relating to trade with colonial groups.
---

@@

One of the most commonly traded types of object in this region were personal ornaments such as glass and stone beads [@Chen2007; @Li2014; @NMTH2005]. Personal adornments in the archaeological record are useful as signal of an individual's status [@Joyce2005; @Scaramelli2005]. For example, the consumption of stone beads in Southeast Asia during Iron Age is often associated with increasing social stratification or socio-political complexity [@Bellina2014; @Carter2016; @Francis2002; @Theunissen2000; @Kenoyer2000]. In this paper, I explore archaeological ornaments from Kiwulan spanning the pre-European contact period, the period of Spanish and Dutch presence, and the period of Chinese presence. I address the question of whether indirect colonial influences on the indigenous populations can be detected through the ornament assemblages. 
# Historical background

The island of Taiwan lies at the junction of mainland China, southeast Asia, and northeast Asia in the Pacific Ocean. The prehistory of Taiwan island could be roughly divided into three major periods, paleolithic (c. 27,000 BP- 5000BP), Neolithic (c. 6500- 2000BP), and Iron age (c. 2000- 400BP) with slightly regional differences in onset of each period and variations in style of artifacts and assemblages (@Liu2011; @Chen2017). It is generally accepted that Taiwan entered historical period since the early 17th century due to the colonial presence of the Spanish and Dutch who played an important role in keeping written records about Taiwan. The Spanish founded Fort San Salvador at Keelung in 1626 and Fort San Domingo in 1629 at Tamsui, and sent missionaries to local indigenous settlements in this region for religious purposes [@Blusse2000, pp. 343]. A report by a Dominican priest in 1632 mentioned that the Taparri, an indigenous tribe from northern Taiwan, exchanged agate beads with other indigenous groups. This form of exchange was widespread and even the Spanish soldiers learned to use agate as bargaining chips for gambling [@LiandWu2006, pp. 132-149]. The use of beads as prestige goods is indicated by their role in bride price payments, and compensation to resolve disputes [@LiandWu2006, pp. 132-149]. Other records mention that the female shamans (majuorbol) in the tribe would use agate beads as magical items in ritual practice for body healing [@Borao2009, pp. 122-151]. Records of an indigenous funeral process document the use of agate beads in ritual contexts, with more agate beads, pottery, and cloth placed into the graves of more influential people to indicate a family’s higher prestige [@LiandWu2006, pp. 153]. These historical accounts indicate that agate beads were already treated as prestige goods in Yilan before the arrival of Europeans.

@@

The chronology of Kiwulan can be divided into two phases represented by a upper component (700-1200 AD) and a lower component (1400-1900 AD) separated by a sterile. This chronology is based on the stratigraphy and a series of 32 radiocarbon dates previously published by @Chen2007. This paper focuses on the upper component because this has a sequence spanning pre-European contact, European presence, and Chinese presence. Previous researchers divided the upper component into six layers spanning from the 14th century to the 19th century with hundred year intervals according to the radiocarbon dates, excavation depth, consistency of contexts, and types of diagnostic porcelains such as blue and white porcelains [@Hsieh2009; @Wang2011]. Due to ambiguities in the previous chronology, I re-examined the excavation records to produce a new chronology for the sampled squares. For each sampled square I separated excavation units into three episodes of time: pre-European period, European period, and Chinese period to indicate the dominant foreign influence and cultural interaction for different time periods.

@@

This study focuses on `r sam_post_hole` ornaments from 40 sampling squares located at the dwelling place of Kiwulan, indicated by aliened postholes with in-situ posts (Figure \@ref(fig:KWL-sam-area)). These postholes were found aligned in a north-south direction in some intervals with construction marks, which were interpreted as the remains of silt house structures. At the north margin of the dwelling place were burials that are mostly oriented in an east-west direction across the sampling area. There are 30 burials in the sampling area account for one third of the total number of burials at Kiwulan. Ornaments are commonly used as grave goods in burials, and the total number including different ornament classes are 3,173. The use of ornaments as body adornments in burials will be discussed to understand the functions of ornaments, but not statistically analysed since the number of ornaments is highly affected by the presence of bead strings, which sometimes contains thousands of beads in an individual burial [@Chen2007]. In addition to ornaments from the dwelling place and burial, there are 27 ornaments found in middens, which to some extent supports that ornaments were treated as high value items at Kiwulan because of small number in the midden contexts compared to high frequency in burial contexts. It is important to analyse the detailed distribution of ornaments from the dwelling place as it can provide key information concerning social inequality based on uneven access to the high-value items, and changes in the spatial pattern of ornament across different time periods.

The ornaments found in the Figure \@ref(fig:plot-all-ornaments-count) shows that the most common ornament are golden beads (n = `r golden_bead_n`), followed by glass beads (n =  `r glass_bead_n`), agate beads (n = `r agate_bead_n`), metal rings (n = `r metal_ring_n`), and metal bells (n = `r metal_bell_n`, see also Figure \@ref(fig:ornaments-image)). The variety of subtypes and their individual frequencies are summarized in Table \@ref(tab:tidy-data-for-table). The chemical composition analysis for glass beads shows high-lead content in the glass, coupled with the winding/folding technique of bead making, suggest a China production tradition[@Cheng2008; @Wang2018]. The style of other ornaments, including glass beads, metal bells, agate beads, are also associate with a Chinese origin [@Chen2007]. There is no clear evidence shows the European origin of beads, however, the large amount of golden beads at Kiwulan might be related to the colonial presence of the Europeans via economic activities because similar beads were found at Luzon, northern Philippines, a part of the trading route of the Spanish between 16-19th century [@Wang2007].

@@

Figure \@ref(fig:plot-all-densities) shows that there is one major core area during the pre-European period, multiple core areas during European period, and a single core during the Chinese period. There are three consistent sub-regions with a core area that shifts over time. The distribution might indicate increase and decrease in the number of different social groups who possessed more ornaments. The multiple groups during the European period might reflect more unequal consumption of ornaments across the site, relative to other periods, or random patterns resulting from bigger sample size. In addition, the generation of core areas might be biased due to small sample sizes, for example, a few ornaments found at one single square during the Chinese period could create an obvious hotspot.

@@

An indirect colonial influence is indicated at Kiwulan by a greater diversity of ornaments types and materials during the European period. The origin of ornaments

This likely results from participation in large scale exchange networks that stimulated the circulation of different ornament classes. However, the frequency of overall ornaments and each subtype declines significantly during the Chinese period. This may be due to a smaller scale of trading network, the overall decline of the indigenous population in Yilan, and adopting the practices of Han Chinese who did not use these kinds of beads in their dress. Since the end of the 18th century, many indigenous people moved southwards to Hualien due to population pressure caused by Han Chinese immigrants [@Chen2007].

The explanations for the increase of ornaments during the European period could be

should be related to participation in global trade networks is not clarified.

On the contrary, a more economical interpretation could be that they are used to vindicate the local customs and resist against European intrusion, rather than the opposite, as the author does. Without using foreign elements (something possible in Kiwulan, where 300+ tobacco pipes have been found, for example) it is difficult to see the influence of global networks in any particular place.
